*Prochlorococcus marinus* responses to light and oxygen

### 2024-02-15 Code/Manuscript_O2.Rmd

*Prochlorococcus* are difficult to culture, as their reduced genome – the smallest of any known oxyphototroph – render them partially dependent upon mutualistic heterotrophic bacteria to  detoxify reactive oxygen species (Moore et al., 2007; Morris et al., 2008). MED4, SS120 and MIT9313 have been successfully cultured in laboratories, and used to show that ecotypic classifications correspond to biochemical differences among strains. CITATONS OUR LAB and OTHERS

The main target of photoinhibition is PSII, in particular its PsbA protein, which along with its homolog PsbD forms a heterodimer binding all redox active components at the photosystem’s reaction center core and is therefore the most vulnerable to redox damage (Barber, 2012). Under light-induced damage, PsbA transcripts accumulate as the organism attempts to replace the damaged protein, however protein levels remain relatively constant, if inoperative. The new copy cannot be incorporated until the damaged one is degraded thus, the degradation of PsbA has been considered the rate-limiting step in recovery from photoinhibition (Kanervo et al., 1993). The proteins believed to be responsible for this crucial step are the FtsH proteases, which have been found localized to the thylakoid membrane of cyanobacteria and higher plants (Sacharz et al., 2015) as well as co-purified with His-tagged PSII (Zak et al., 2001). Furthermore, mutants lacking FtsH have increased sensitivity to photoinactivation (Mann et al., 2000; Adam et al., 2005). 
Originally identified in Escherichia coli, (Tomoyasu et al., 1993) FtsH is a membrane-bound, ATP, zinc metalloprotease that extracts and degrades membrane-bound proteins for degradation (Chiba et al., 2002). FtsH are embedded in membranes via two N-terminal transmembrane domains connected by a small hydrophilic loop in the thylakoid lumen (Figure 9A). The main hydrophilic portion sits in the bacterial cytoplasm and contains an ATP-ase Associated with various cellular Activities (AAA) domain as well as a zinc-binding active site (Yoshioka-Nishimura and Yamamoto, 2014) (Figure 10a). Chloroplast FtsH were first identified when immunoblot analysis with antibodies against E. coli FtsH cross reacted with thylakoid membrane proteins in spinach (Lindahl et al., 1996). The chloroplast genome of higher plants can encode up to 8 FtsH homologs (Adam et al., 2005). Cyanobacterial genomes encode only 4 homologs. In the model cyanobacterium Synechocystis sp. PCC 6803 (Synechocystis) these are encoded by the slr1390, slr0228, slr1604 and sll1463 genes, which will be used to identify their respective proteins for clarity.  Slr0228 and sll1463 are found localized to patches in the thylakoid membrane (Sacharz et al., 2015) and it has been shown that sll1463 accumulation is dependent upon slr0228 (Pisareva et al., 2007). Furthermore, slr0228 has been shown to be indispensable for PSII repair and PSI biogenesis but intriguingly is not required for cell viability, nor is sll1463 (Mann et al., 2000). In contrast, slr1390 and slr1604 are found in both plasma and thylakoid membranes (Pisareva et al., 2007; Sacharz et al., 2015) and are essential for cell viability (Mann et al., 2000).  FtsH proteases have been shown to associate in homo- and hetero-oligomers (Yu et al., 2004). In Synechocystis, slr1390 and slr1604 associate in hetero-oligomers (Sacharz et al., 2015). Slr0228 and slr1604 form a hetero-hexameric ring in the thylakoid membrane with a central pore through which proteins are translocated as they are digested and it is this complex that is responsible for PSII repair (Adam et al., 2005; Boehm et al., 2012) (Figure 10b). Nixon et al. (2010) have proposed a model for the PSII repair process in which destabilization of PsbA, resulting in either unusual exposure or post-translational modification of PsbA’s N-terminus (Komenda et al., 2007), exposes an N-terminal “docking site” for FtsH to initiate  degradation (Figure 11).

Representatives of LL clades of *Prochlorococcus* suffer photoinhibition at irradiance levels preferred by their HL relatives. The main target of photoinhibition is PSII, in particular the PsbA protein, which along with its homolog PsbD forms a heterodimer binding all redox active components at the photosystem’s reaction center core and is therefore the most vulnerable to redox damage (Barber, 2012). Under light-induced damage, new PsbA cannot be incorporated until the damaged PsbA is removed thus, degradation of PsbA is a rate-limiting step in recovery from photoinhibition (Kanervo et al., 1993). PsbA degradation is mediated largely by FtsH proteases (Mann et al., 2000; Adam et al., 2005), localized to the thylakoid membrane (Sacharz et al., 2015) and PSII (Zak et al., 2001). FtsH is a membrane-bound, ATP, zinc metalloprotease that extracts and degrades membrane-bound proteins for degradation (Chiba et al., 2002) (Yoshioka-Nishimura and Yamamoto, 2014). In the model cyanobacterium *Synechocystis* sp. PCC 6803 4 isoforms of FtsH proteins are encoded by the slr1390, slr0228, slr1604 and sll1463 genes.  Slr0228 and sll1463 are found localized to patches in the thylakoid membrane (Sacharz et al., 2015) and it has been shown that sll1463 accumulation is dependent upon slr0228 (Pisareva et al., 2007). Furthermore, slr0228 has been shown to be indispensable for PSII repair and PSI biogenesis but intriguingly is not required for cell viability, nor is sll1463 (Mann et al., 2000). In contrast, slr1390 and slr1604 are found in both plasma and thylakoid membranes (Pisareva et al., 2007; Sacharz et al., 2015) and are essential for cell viability (Mann et al., 2000).  FtsH proteases have been shown to associate in homo- and hetero-oligomers (Yu et al., 2004). In Synechocystis, slr1390 and slr1604 associate in hetero-oligomers (Sacharz et al., 2015). Slr0228 and slr1604 form a hetero-hexameric ring in the thylakoid membrane with a central pore through which proteins are translocated as they are digested and it is this complex that is responsible for PSII repair (Adam et al., 2005; Boehm et al., 2012) (Figure 10b). Nixon et al. (2010) have proposed a model for the PSII repair process in which destabilization of PsbA, resulting in either unusual exposure or post-translational modification of PsbA’s N-terminus (Komenda et al., 2007), exposes an N-terminal “docking site” for FtsH to initiate  degradation (Figure 11).

*Prochlorococcus* genomes encode 4 FtsH proteins, henceforth referred to as FtsH1-4. Bonisteel et al. (2018) found that these were homologs of *Synechocystis* FtsH (Table 1).  Upon a shift to higher light MED4 upregulated expression of FtsH1 and FtsH2, homologous to *Synechocystis* slr0228 and slr1604, implicated in PSII repair. In contrast representative LL strain MIT9313 shows less expression of the FtsH proteases, and thus has fewer FtsH serving each photosystem,  MIT9313 expressed primarily FtsH3 and expression did not change in response to light stress. Bonisteel concluded with the hypothesis that LL Prochlorococcus may be allocating their energy resources to other processes that would advantage life under low light conditions.

@@

This correlates with the above research on photosystem composition in laboratory cultivated HL and LL Prochlorococcus. HL strains are found primarily in oligotrophic surface waters where light is abundant, thus allocate more of their energy to maintaining their photosynthetic apparatuses and can obtain energy from the reduction of sugars. In contrast, LL strains live under eponymically low irradiances and may depend on environmental nutrients and cyclic electron flow to maintain energy homeostasis.

>>>>

*Prochlorococcus* are difficult to culture, as their reduced genomes – the smallest of any known oxyphototroph – render them partially dependent upon mutualistic heterotrophic bacteria to  detoxify reactive oxygen species (Moore et al., 2007; Morris et al., 2008). MED4, SS120 and MIT9313 have been successfully cultured in laboratories, and used to show that ecotypic classifications correspond to biochemical differences among strains. CITATONS OUR LAB and OTHERS.

Under full atmospheric [O~2~], LL clades of *Prochlorococcus* are restricted to growth under low light, in part because they suffer photoinhibition of Photosystem II (PSII) at irradiances preferred by their HL relatives. The main target of photoinhibition is the PsbA protein, which along with its homolog PsbD forms a heterodimer binding all redox active components of the PSII reaction center core (Barber, 2012). Under light-induced damage, new PsbA cannot be incorporated until the damaged PsbA is removed.  Thus, degradation of PsbA, mediated largely by FtsH proteases (Mann et al., 2000; Adam et al., 2005), is a rate-limiting step in recovery from photoinhibition (Kanervo et al., 1993). FtsH is a membrane-bound (Sacharz et al., 2015)(Zak et al., 2001), ATP, zinc metalloprotease that extracts and degrades membrane-bound proteins for degradation (Chiba et al., 2002) (Yoshioka-Nishimura and Yamamoto, 2014). In the model cyanobacterium *Synechocystis* sp. PCC 6803 4 isoforms of FtsH proteins are encoded by the slr1390, slr0228, slr1604 and sll1463 genes.  In *Synechocystis*, Slr0228 and slr1604 form a hetero-hexameric ring in the thylakoid membrane with a central pore through which proteins are translocated as they are digested, a process key for PSII repair (Adam et al., 2005; Boehm et al., 2012) (Nixon et al. (2010)) (Komenda et al., 2007), when photoinactivation exposes an N-terminal “docking site” for FtsH to initiate degradation. In contrast, slr1390 and slr1604 are found in both plasma and thylakoid membranes (Pisareva et al., 2007; Sacharz et al., 2015), are essential for cell viability (Mann et al., 2000), and implicated in PSI biogenesis (AMANDA REF from ALGATECH).

*Prochlorococcus* genomes encode 4 FtsH proteins, henceforth referred to as FtsH1-4, homologous to the *Synechocystis* FtsH isofoms, and with presumably parallel functions (Table 1).  Upon a shift to higher light HLIII MED4 upregulated expression of FtsH1 and FtsH2 (Bonisteel et al. (2018)), homologs to *Synechocystis* slr0228 and slr1604, implicated in PSII repair. In contrast representative LLIII strain MIT9313 shows less overall expression of the FtsH proteases, and thus has fewer FtsH serving each photosystem.  Furthermore MIT9313 expressed primarily FtsH3 and expression did not increase in response to light stress. Through adaptation to steady low light, LLIII **Prochlorococcus* allocate resources to other processes.

@@
